New cardinality estimation algorithms for HyperLogLog sketches

### 2016-06-11 paper/paper.tex
paper
added GitHub link to paper, estimation error section for corrected raw
estimator, replaced "linear probabilistic counting" by "linear
counting", some other minor changes

\title{New cardinality estimation algorithms for HyperLogLog sketches}
\subtitle{Draft version}
\author{Otmar Ertl}

@@

Note that a $(\symPrecision, 0)$-HyperLogLog sketch corresponds to a bit array as used by linear probabilistic counting \cite{Whang1990}. In this case each register value is represented by a single bit. Hence, linear probabilistic counting can be regarded as a special case of the HyperLogLog algorithm ($\symRegRange = 0$).

@@

In the history of the HyperLogLog algorithm different averaging techniques have been proposed. First, there was the LogLog algorithm using the geometric mean and the SuperLogLog algorithm that enhanced the estimate by truncating the largest register values before applying the geometric mean \cite{Durand2003}. Finally, the harmonic mean was found to give even better estimates as it is inherently less sensitive to outliers. The result is the so-called raw HyperLogLog estimate and given by
\begin{equation}
\label{equ:intermediate_estimate}
\symCardinalityOriginalEstimate
= 
\frac{\symAlpha_\symNumReg \symNumReg^2}{\sum_{\symIndexJ=1}^{\symNumReg}2^{-\symRegVal_\symIndexJ}}
= 
\frac{\symAlpha_\symNumReg \symNumReg^2}{\sum_{\symRegVal=0}^{\symRegRange+1}\symCount_\symRegVal 2^{-\symRegVal}}
\end{equation}
where $\symAlpha_\symNumReg$  is a correction factor that is fixed for a given number of registers $\symNumReg$ and which was derived to be \cite{Flajolet2007}
\begin{equation}
\symAlpha_\symNumReg := \left(
\symNumReg
\int_0^\infty
\left(
\log_2\left(
\frac{2+u}{1+u}
\right)
\right)^\symNumReg
du
\right)^{-1}.
\end{equation}
Numerical approximations of $\symAlpha_\symNumReg$ for various values of $\symNumReg$ have been listed in \cite{Flajolet2007}. These approximations are used in many HyperLogLog implementations. However, since the published constants have been rounded to 4 significant digits, these approximations even introduce some bias for very high precisions $\symPrecision\geq 20$. For HyperLogLog sketches that are used in practice with 256 or more registers ($\symPrecision\geq 8$), it is completely sufficient to use 
\begin{equation}
\label{equ:alpha_inf}
\symAlpha_\infty := \lim_{\symNumReg\rightarrow\infty} \symAlpha_\symNumReg = \frac{1}{2\log 2} \approx 0.7213475,
\end{equation}
as approximation for $\symAlpha_\symNumReg$ in \eqref{equ:intermediate_estimate}, because the additional bias is negligible if compared to the estimation error.

@@

As mentioned in \cref{sec:data_element_insertion}, a HyperLogLog sketch with parameters $(\symPrecision, \symRegRange)$ can be turned into a $(\symPrecision, 0)$-HyperLogLog sketch. Since $\symRegRange=0$ corresponds to linear probabilistic counting and the transformed HyperLogLog sketch corresponds to a bitset with $\symCount_0$ zeros, we can apply the corresponding cardinality estimator \cite{Whang1990}
\begin{equation}
\label{equ:small_range_estimate}
\symCardinalityEstimate_\text{small} = \symNumReg \log\left(
\frac{\symNumReg}{\symCount_0}
\right).
\end{equation}
The corresponding relative estimation error as depicted in \cref{fig:small_range_estimate} shows that this estimator is convenient for small cardinalities. It was proposed to use this small-range estimator as long as $\symCardinalityOriginalEstimate \leq \frac{5}{2}\symNumReg$ where the factor $\frac{5}{2}$ was empirically determined \cite{Flajolet2007}.

@@

\begin{figure}
\centering
\includesvg[width=1\textwidth]{small_range_estimate}
\caption{The distribution of the relative estimation error for the linear probabilistic counting estimator which can be used for small cardinalities.}
\label{fig:small_range_estimate}
\end{figure}

@@

\subsection{Derivation of the raw estimator}
In order to better understand why the raw estimator fails for small and large cardinalities, we start with a brief derivation that does not require complex analysis.

@@

Note that, $2\symSmallCorrectionFunc(\symX) + \symLargeCorrectionFunc(\symX) = -\symPowerSeriesFunc(\log_2(-\log(\symX)))/\left(\log(2)\log(\symX)\right)$. Therefore, we get for the special case of linear probabilistic counting with $\symRegRange=0$
\begin{equation}
\symPoissonRateEstimate 
= 
\frac{\symNumReg}{\log(2)\left(
2\symSmallCorrectionFunc(\symCount_0/\symNumReg)
+
\symLargeCorrectionFunc(\symCount_0/\symNumReg)
\right)}
=
\frac{\symNumReg\log\left(\frac{\symNumReg}{\symCount_0}\right)}{\symPowerSeriesFunc\left(\log_2\left(\log\left(\frac{\symNumReg}{\symCount_0}\right)\right)\right)}
\approx
\symNumReg\log\left(\frac{\symNumReg}{\symCount_0}\right).
\end{equation}
almost the same estimator as given in \eqref{equ:small_range_estimate}. Here we used the fact that $\symPowerSeriesFunc$ can be well approximated by 1.
\subsection{Corrected raw estimation algorithm}
The corrected raw estimator \eqref{equ:correctedestimator} leads to a new cardinality estimation algorithm for HyperLogLog sketches. \cref{alg:corrected_raw_estimation} demonstrates the numerical evaluation of the estimator. The values of functions $\symSmallCorrectionFunc$ and $\symLargeCorrectionFunc$ can be either calculated on-demand or pre-calculated. The possible value range for $\symCount_0$ and $\symCount_{\symRegRange+1}$ is $[0, \symNumReg]$. Therefore, if performance matters, it is possible to precalculate the function values for all possible arguments and keep them in lookup tables of size $\symNumReg+1$.

@@

see \cref{fig:raw_corrected_estimation_error_8_24}, \cref{fig:raw_corrected_estimation_error_12_20},
\cref{fig:raw_corrected_estimation_error_16_16},
\cref{fig:raw_corrected_estimation_error_22_10},

@@

@@

Note that \eqref{equ:funcdef} can be solved analytically for the special case $\symRegRange=0$ which corresponds to the already mentioned linear probabilistic counting algorithm. In this case, the maximum likelihood method under the Poisson model directly leads to the estimator presented in \cite{Whang1990} and which was also used for small range estimation \eqref{equ:small_range_estimate}. Despite the assumption of a Poisson model, it is a very good approximation of the optimal martingale estimator presented in \cite{Ting2014}. Due to this fact we could expect that maximum likelihood estimation under the Poisson model also works very well for the more  general HyperLogLog case.

@@

\cref{fig:estimation_error_12_20} shows the distribution of the relative error of the estimated cardinality using \cref{alg:estimate_ml} compared to the true cardinality for $\symPrecision=12$ and $\symRegRange=20$. As the median shows, the error is essentially unbiased over the entire cardinality range except for very small values. The new approach is able to accurately estimate cardinalities up to 4 billions ($\approx 2^{\symPrecision+\symRegRange}$) which is about an order of magnitude larger than the operating range upper bound of the raw estimate (\cref{fig:raw_estimate}) or the original method (\cref{fig:original_estimate}).

\begin{figure}
\centering
\includesvg[width=1\textwidth]{max_likelihood_estimate_12_20}
\caption{Relative estimation error as a function of the true cardinality for a HyperLogLog sketch with parameters $\symPrecision = 12$ and $\symRegRange=20$.}
\label{fig:estimation_error_12_20}
\end{figure}

The new estimation algorithm also works well for other HyperLogLog configurations. First we considered the original HyperLogLog algorithm that used a 32-bit hash function ($\symPrecision + \symRegRange = 32$). The relative estimation error for precisions $\symPrecision=8$, $\symPrecision=16$, $\symPrecision=22$ are shown in \cref{fig:estimation_error_8_24}, \cref{fig:estimation_error_16_16}, and \cref{fig:estimation_error_22_10}, respectively. As expected, since  $\symPrecision + \symRegRange = 32$ is kept constant, the operating range remains more or less the same, while the relative error decreases with increasing precision. Again, the new algorithm gives inherently unbiased estimates.

@@

As proposed in \cite{Heule2013}, the operating range can be extended by
replacing the 32-bit hash function by a 64-bit hash function. \cref{fig:estimation_error_12_52} shows the relative error for such a HyperLogLog configuration with precision  $\symPrecision=12$. In this case, in order to use all the 64 bits of the hash value, $\symRegRange$ must be chosen to be equal to $64 - \symPrecision = 52$. As a consequence, in order to represent the maximum possible register value $\symRegRange+1 = 53$, 6 bits are needed for each register. The doubled hash value size shifts the maximum trackable cardinality value towards $2^{64}$. As \cref{fig:estimation_error_12_52} shows, when compared to the 32-bit hash value case given in \cref{fig:estimation_error_12_20}, the estimation error remains constant over the entire simulated cardinality range up to 50 billions.

@@

We also evaluated the case $\symPrecision = 12$ and $\symRegRange=14$, which is interesting, because the register values are limited to the range $[0, 15]$. As a consequence, 4 bits are sufficient for representing a single register value. This allows two registers to share a single byte, which is beneficial from a performance perspective. Nevertheless, this configuration still allows the estimation of cardinalities up to 100 millions as shown in \cref{fig:estimation_error_12_14}, which could be enough for many applications.

@@

If $\symHelper$ is not calculated exactly but approximated by an approximation $\symHelperApprox$ with maximum relative error $\symError_\symHelper\ll 1$

>>>>

\title{New cardinality estimation algorithms for HyperLogLog sketches}
\subtitle{Draft version, \url{https://github.com/oertl/hyperloglog-sketch-estimation-paper}}
\author{Otmar Ertl}

@@

Note that a $(\symPrecision, 0)$-HyperLogLog sketch corresponds to a bit array as used by linear counting \cite{Whang1990}. In this case each register value is represented by a single bit. Hence, linear counting can be regarded as a special case of the HyperLogLog algorithm ($\symRegRange = 0$).

@@

In the history of the HyperLogLog algorithm different averaging techniques have been proposed. First, there was the LogLog algorithm using the geometric mean and the SuperLogLog algorithm that enhanced the estimate by truncating the largest register values before applying the geometric mean \cite{Durand2003}. Finally, the harmonic mean was found to give even better estimates as it is inherently less sensitive to outliers. The result is the so-called raw HyperLogLog estimate and given by
\begin{equation}
\label{equ:intermediate_estimate}
\symCardinalityOriginalEstimate
= 
\frac{\symAlpha_\symNumReg \symNumReg^2}{\sum_{\symIndexJ=1}^{\symNumReg}2^{-\symRegVal_\symIndexJ}}
= 
\frac{\symAlpha_\symNumReg \symNumReg^2}{\sum_{\symRegVal=0}^{\symRegRange+1}\symCount_\symRegVal 2^{-\symRegVal}}
\end{equation}
where $\symAlpha_\symNumReg$  is a correction factor that is fixed for a given number of registers $\symNumReg$ and which was derived to be \cite{Flajolet2007}
\begin{equation}
\symAlpha_\symNumReg := \left(
\symNumReg
\int_0^\infty
\left(
\log_2\left(
\frac{2+u}{1+u}
\right)
\right)^\symNumReg
du
\right)^{-1}.
\end{equation}
Numerical approximations of $\symAlpha_\symNumReg$ for various values of $\symNumReg$ have been listed in \cite{Flajolet2007}. These approximations are used in many HyperLogLog implementations. However, since the published constants have been rounded to 4 significant digits, these approximations even introduce some bias for very high precisions. For HyperLogLog sketches that are used in practice with 256 or more registers ($\symPrecision\geq 8$), it is completely sufficient to use 
\begin{equation}
\label{equ:alpha_inf}
\symAlpha_\infty := \lim_{\symNumReg\rightarrow\infty} \symAlpha_\symNumReg = \frac{1}{2\log 2} \approx 0.7213475,
\end{equation}
as approximation for $\symAlpha_\symNumReg$ in \eqref{equ:intermediate_estimate}, because the additional bias is negligible if compared to the estimation error.

@@

As mentioned in \cref{sec:data_element_insertion}, a HyperLogLog sketch with parameters $(\symPrecision, \symRegRange)$ can be turned into a $(\symPrecision, 0)$-HyperLogLog sketch. Since $\symRegRange=0$ corresponds to linear counting and the transformed HyperLogLog sketch corresponds to a bitset with $\symCount_0$ zeros, the linear counting cardinality estimator \cite{Whang1990} can be applied here
\begin{equation}
\label{equ:small_range_estimate}
\symCardinalityEstimate_\text{small} = \symNumReg \log\left(
\frac{\symNumReg}{\symCount_0}
\right).
\end{equation}
The corresponding relative estimation error as depicted in \cref{fig:small_range_estimate} shows that this estimator is convenient for small cardinalities. It was proposed to use this small-range estimator as long as $\symCardinalityOriginalEstimate \leq \frac{5}{2}\symNumReg$ where the factor $\frac{5}{2}$ was empirically determined \cite{Flajolet2007}.

@@

\begin{figure}
\centering
\includesvg[width=1\textwidth]{small_range_estimate}
\caption{The distribution of the relative estimation error for the linear counting estimator which can be used for small cardinalities.}
\label{fig:small_range_estimate}
\end{figure}

@@

\subsection{Derivation of the raw estimator}
\label{sec:derivation_raw_estimator}
In order to better understand why the raw estimator fails for small and large cardinalities, we start with a brief derivation that does not require complex analysis.

@@

Using the identity $2\symSmallCorrectionFunc(\symX) + \symLargeCorrectionFunc(\symX) = -\symPowerSeriesFunc(\log_2(-\log(\symX)))/\left(\log(2)\log(\symX)\right)$, we get for the linear counting case with $\symRegRange=0$
\begin{equation}
\symPoissonRateEstimate 
= 
\frac{\symNumReg}{\log(2)\left(
2\symSmallCorrectionFunc(\symCount_0/\symNumReg)
+
\symLargeCorrectionFunc(\symCount_0/\symNumReg)
\right)}
=
\frac{\symNumReg\log\left(\frac{\symNumReg}{\symCount_0}\right)}{\symPowerSeriesFunc\left(\log_2\left(\log\left(\frac{\symNumReg}{\symCount_0}\right)\right)\right)}
\approx
\symNumReg\log\left(\frac{\symNumReg}{\symCount_0}\right),
\end{equation}
which is almost identical to the linear counting estimator used for the small range correction \eqref{equ:small_range_estimate}. Here we used the fact that $\symPowerSeriesFunc$ can be well approximated by 1.
\subsection{Corrected raw estimation algorithm}
The corrected raw estimator \eqref{equ:correctedestimator} leads to a new cardinality estimation algorithm for HyperLogLog sketches. \cref{alg:corrected_raw_estimation} demonstrates the numerical evaluation of the estimator. The values of functions $\symSmallCorrectionFunc$ and $\symLargeCorrectionFunc$ can be either calculated on-demand or pre-calculated. The possible value range for $\symCount_0$ and $\symCount_{\symRegRange+1}$ is $[0, \symNumReg]$. Therefore, if performance matters, it is possible to precalculate the function values for all possible arguments and keep them in lookup tables of size $\symNumReg+1$.

@@

\cref{fig:raw_corrected_estimation_error_12_20} shows the distribution of the relative error of the estimated cardinality using \cref{alg:corrected_raw_estimation} compared to the true cardinality for $\symPrecision=12$ and $\symRegRange=20$. As the mean shows, the error is unbiased over the entire cardinality range. The new approach is able to accurately estimate cardinalities up to 4 billions ($\approx 2^{\symPrecision+\symRegRange}$) which is about an order of magnitude larger than the operating range upper bound of the raw estimator (\cref{fig:raw_estimate}) or the original method (\cref{fig:original_estimate}).

The new estimation algorithm also works well for other HyperLogLog configurations. First we considered the original HyperLogLog algorithm that used a 32-bit hash function ($\symPrecision + \symRegRange = 32$). The relative estimation error for precisions $\symPrecision=8$, $\symPrecision=16$, $\symPrecision=22$ are shown in \cref{fig:raw_corrected_estimation_error_8_24}, \cref{fig:raw_corrected_estimation_error_16_16}, and \cref{fig:raw_corrected_estimation_error_22_10}, respectively. As expected, since  $\symPrecision + \symRegRange = 32$ is kept constant, the operating range remains more or less the same, while the relative error decreases with increasing precision. Again, the new algorithm gives essentially unbiased estimates. Only for very large precisions, an oscillating bias becomes apparent (compare \cref{fig:raw_corrected_estimation_error_22_10}), that is caused by approximating the periodic function $\symPowerSeriesFunc$ by a constant (see \cref{sec:derivation_raw_estimator}).

@@

As proposed in \cite{Heule2013}, the operating range can be extended by
replacing the 32-bit hash function by a 64-bit hash function. \cref{fig:raw_corrected_estimation_error_12_52} shows the relative error for such a HyperLogLog configuration with precision  $\symPrecision=12$. In this case, in order to use all the 64 bits of the hash value, $\symRegRange$ must be chosen to be equal to $64 - \symPrecision = 52$. As a consequence, in order to represent the maximum possible register value $\symRegRange+1 = 53$, 6 bits are needed for each register. The doubled hash value size shifts the maximum trackable cardinality value towards $2^{64}$. As \cref{fig:raw_corrected_estimation_error_12_52} shows, when compared to the 32-bit hash value case given in \cref{fig:raw_corrected_estimation_error_12_20}, the estimation error remains constant over the entire simulated cardinality range up to 50 billions.

\begin{figure}
\centering
\includesvg[width=1\textwidth]{raw_corrected_estimate_12_52}
\caption{Relative estimation error as a function of the true cardinality for a HyperLogLog sketch with parameters $\symPrecision = 12$ and $\symRegRange=52$.}
\label{fig:raw_corrected_estimation_error_12_52}
\end{figure}

We also evaluated the case $\symPrecision = 12$ and $\symRegRange=14$, which is interesting, because the register values are limited to the range $[0, 15]$. As a consequence, 4 bits are sufficient for representing a single register value. This allows two registers to share a single byte, which is beneficial from a performance perspective. Nevertheless, this configuration still allows the estimation of cardinalities up to 100 millions as shown in \cref{fig:raw_corrected_estimation_error_12_14}, which could be enough for many applications.

\begin{figure}
\centering
\includesvg[width=1\textwidth]{raw_corrected_estimate_12_14}
\caption{Relative estimation error as a function of the true cardinality for a HyperLogLog sketch with parameters $\symPrecision = 12$ and $\symRegRange=14$.}
\label{fig:raw_corrected_estimation_error_12_14}
\end{figure}

@@

Note that \eqref{equ:funcdef} can be solved analytically for the special case $\symRegRange=0$ which corresponds to the already mentioned linear counting algorithm. In this case, the maximum likelihood method under the Poisson model directly leads to the estimator presented in \cite{Whang1990} and which was also used for small range estimation \eqref{equ:small_range_estimate}. Despite the assumption of a Poisson model, it is a very good approximation of the optimal martingale estimator presented in \cite{Ting2014}. Due to this fact we could expect that maximum likelihood estimation under the Poisson model also works very well for the more  general HyperLogLog case.

@@

see \cref{fig:estimation_error_12_20}, \cref{fig:estimation_error_8_24}, \cref{fig:estimation_error_16_16},\cref{fig:estimation_error_22_10},\cref{fig:estimation_error_12_52},\cref{fig:estimation_error_12_14}
\begin{figure}
\centering
\includesvg[width=1\textwidth]{max_likelihood_estimate_12_20}
\caption{Relative estimation error as a function of the true cardinality for a HyperLogLog sketch with parameters $\symPrecision = 12$ and $\symRegRange=20$.}
\label{fig:estimation_error_12_20}
\end{figure}

@@

@@

@@

If $\symHelper$ is not calculated exactly but approximated by $\symHelperApprox$ with maximum relative error $\symError_\symHelper\ll 1$
